---
title: "Quickstart Guide: zzrenvcheck"
author: "Ronald (Ryy) G. Thomas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart Guide: zzrenvcheck}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

**zzrenvcheck** validates that all R packages used in your code are properly
declared in DESCRIPTION and locked in renv.lock, maintaining reproducible
environments for collaborative research.

## Installation

```{r install}
# Install from GitHub
remotes::install_github("rgt47/zzrenvcheck")
```

```{r load}
library(zzrenvcheck)
```

## Basic Usage

### Check Your Project

```{r check}
# Validate current project
check_packages()
```

This scans your code for package usage and compares against DESCRIPTION
and renv.lock.

### Auto-Fix Missing Packages

```{r fix}
# Add missing packages automatically
fix_packages()
```

This queries CRAN/Bioconductor to find and add missing packages.

### Generate Status Report

```{r report}
# Get detailed status
report <- report_packages()
print(report)

# Output:
#   package         in_code in_description in_renv_lock status
#   dplyr           TRUE    FALSE          FALSE        missing_description
#   ggplot2         TRUE    TRUE           TRUE         ok
```

## Key Features

### Package Detection Patterns

The package detects packages from:

- `library(dplyr)` - Direct library calls
- `require(ggplot2)` - Conditional loading
- `tidyr::pivot_longer()` - Namespace calls
- `#' @importFrom dplyr filter` - Roxygen imports

### Smart Filtering

19 filters avoid false positives:

- Base R packages (base, utils, stats)
- Placeholder names (myproject, package, foo)
- Generic words (my, your, file, path)
- Invalid package names

### Validation Modes

```{r modes}
# Standard mode (R/, scripts/, analysis/)
check_packages(strict = FALSE)

# Strict mode (includes tests/, vignettes/, inst/)
check_packages(strict = TRUE)
```

## Sync to Code

Make code the single source of truth:

```{r sync}
# Sync DESCRIPTION and renv.lock to match code
sync_packages()

# Preview changes first
sync_packages(dry_run = TRUE)
```

This adds missing packages and removes unused ones.

## Package Source Validation

Check if packages exist on CRAN, Bioconductor, or GitHub:

```{r validate}
# Single package
is_installable("dplyr")
# Returns: list(installable = TRUE, source = "CRAN", package = "dplyr")

# Batch validation
check_installable(c("dplyr", "DESeq2", "NonExistent"))
```

## Workflow Integration

### Pre-Commit

```bash
# In pre-commit hook
Rscript -e 'zzrenvcheck::check_packages(auto_fix = FALSE)'
```

### CI/CD Pipeline

```yaml
# .github/workflows/check.yaml
- name: Check dependencies
  run: Rscript -e 'zzrenvcheck::check_packages()'
```

### Makefile Integration

```makefile
check-renv:
    Rscript -e 'zzrenvcheck::check_packages()'
```

## Standalone Shell Script

For environments without R:

```bash
# Install shell script version
cd zzrenvcheck
./install.sh --prefix ~/bin

# Use anywhere
zzrenvcheck --fix --strict --verbose
```

Features:

- No R installation required
- Uses grep, sed, awk, jq, curl
- Perfect for CI/CD pipelines

## Function Reference

| Function | Purpose |
|:---------|:--------|
| `check_packages()` | Main validation |
| `fix_packages()` | Auto-fix wrapper |
| `sync_packages()` | Sync to code |
| `report_packages()` | Status report |
| `clean_description()` | Remove unused |
| `is_installable()` | Check CRAN/Bioconductor |
| `extract_code_packages()` | Scan code |

## Common Workflows

### Fix Missing Packages

```{r workflow-fix}
# Check and fix
result <- check_packages(auto_fix = TRUE)

# Then commit
system("git add DESCRIPTION renv.lock")
system('git commit -m "Add missing packages"')
```

### Cleanup Unused Packages

```{r workflow-cleanup}
# Remove packages no longer in code
clean_description()
sync_packages()
```

### Programmatic Use

```{r workflow-programmatic}
result <- check_packages(auto_fix = FALSE)

if (result$status == "fail") {
  cat("Missing packages:\n")
  print(result$missing_in_description)
}
```

## Next Steps

- `?check_packages` - Main function documentation
- `?sync_packages` - Sync workflow details
- GitHub: https://github.com/rgt47/zzrenvcheck
